Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-07-07 joint_inf.tex
Still working on understanding bounds

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
%EM Is it indicated anywhere how the internal nodes are numbered? I would assume that 1 would be connected to leaf 1. If that's true please indicate here.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{2\}$ and internal node $\{2\}$ connected to leaves $\{3\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

%das: did this on paper, but can write it out.
\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
%EM Yeah, this next statement feels like it could use some justification, and or maybe you could skip to the following statement.
Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities the Farris tree topology.
For site split $\emptyset$,
\begin{align*}
    P(\siteSplit=\emptyset|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    P(\siteSplit=\{1\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    P(\siteSplit=\{3\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
$$
\left\{\begin{array}{l}
        (1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2) \\
        (1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2) \\
        (1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2) \\
        (1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2)
\end{array}\right.
$$
for site split $\{1\}$ and
$$
\left\{\begin{array}{l}
        (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2) \\
        (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2) \\
        (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2) \\
        (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
\end{array}\right.
$$
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
By symmetry exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
A similar property holds for $y_1$ and $y_2$.
The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

\begin{table}
%EM I wonder why you use primes on all variables here but not in tab:likelihoods.
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')(1-y')$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2\}\}$&
    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    %EM Shouldn't this be \{1\}\\ ? Or perhaps the ordering of internal nodes is not what I thought?
                    \{1\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1+w')(1-y')^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2\}\}$&
                $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,2,3\}$&
                $\{1,2\}$&
                $(1+x')^2   (1+w')(1+y')(1-y')$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x')(1-x')(1+w')(1+y')^2\\
                        (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')^2   (1+w')(1+y')(1-y')\\
                    (1+x')(1-x')(1-w')(1+y')^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1+x')^2   (1-w')(1+y')^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
                    (1+x')^2   (1+w')(1-y')^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1-x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1-w')(1-y')^2
                    \end{array}$\\
     $\{1,2,3\}$       &
    $\left\{\begin{array}{l}
                    \{2\} \\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1-w')(1+y')^2 \\
                    (1+x')^2   (1+w')(1+y')(1-y')
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state partitions.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Appendix for full calculations.}
\label{tab:likelihoods}
\end{table}

@@

%EM I'd suggest p_\emptyset for this particular one, but perhaps that will throw off your spacing below.

@@

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
$$
a_{1}' = a_{1}-2p_{13},
$$
$$
a_{2}' = a_{2}+2p_{13},
$$
$$
a_{3}' = a_{3}-2p_{13},
$$
$$
a_{4}' = a_{4}+2p_{13}.
$$
Then,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
%EM Sorry to be slow, but could you provide a little more detail on why this follows? Is there some convexity of entropy argument going on here?
%Not sure which part is confusing, but:
    %1) a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
    %The below equation follows by rewriting 1+x and 1-x for the above x.
    %2) a_1' and a_2' are just defined by rewriting the likelihood. There are now two fewer log(1+x) terms and two more log(1-x) terms since we replaced a (1+x)^2 in the likelihood with a (1-x)^2. That's why we have a -2p_{13} and a +2p_{13} respectively.
    %Happy to clarify somewhere in the text but not sure what part would most need exposition.
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
\end{align*}
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.

>>>>

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments is
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
\begin{align*}
    \mathcal{P}(\siteSplitSet) &= \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\} \\
& := \{\siteSplit_1, \ldots, \siteSplit_8\}.
\end{align*}
Since we identify character complements, we do not consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
the symmetry of the binary character model allowing us to focus only on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
Internal node $\{1\}$ is the node connected to leaves $\{1\}$ and $\{2\}$ and internal node $\{2\}$ connected to leaves $\{3\}$ and $\{4\}$.
The mapping from characters to splits in this case will depend on the characters at the tips and the ancestral states.
For example, we take both $\patternToSplit(0000)=\emptyset$ and $\patternToSplit(1111)=\emptyset$.
Similarly, we have $\ancestralToSplit(0000, 00) = \emptyset$ and $\ancestralToSplit(1111, 11)=\emptyset$, needing to take the complement of all the characters present on the tree to identify splits.
We cannot identify complements for ancestral states since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
%EM complete to introduce: We now note that exchanging $x_1$ and $x_2$, or $y_1$ or $y_2$, does not change ...
%EM please be a little more specific than just $ell$, because there are lots of variants-- e.g. likelihood with what data?
Using the Hadamard transform approach outlined in 8.6 of Semple and Steel \cite{Semple2003-em}, we calculate the generating probabilities on the Farris tree topology.
For site split $\emptyset$,
\begin{align*}
    P(\siteSplit=\emptyset|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  x_1y_1w + x_1y_2w + y_1x_2w + x_2y_2w + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1y_1 + x_1y_2 + y_1x_2 + x_2y_2] + x_1y_1x_2y_2) \\
                                              & = \frac{1}{8} \cdot (1 + x_1x_2 +  y_1y_2 +  w[x_1 + x_2][y_1 + y_2] + x_1y_1x_2y_2).
\end{align*}
and this probability is unchanged when $x_1$ is exchanged with $x_2$ and $y_1$ is exchanged with $y_2$.
All other generating probabilities will differ only in the signs of each term.
For example, for site split $\{1\}$ we have
\begin{align*}
    P(\siteSplit=\{1\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[-x_1 + x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
and for site split $\{3\}$ we have
\begin{align*}
    P(\siteSplit=\{3\}|\tau_1, \tilde{t}) & = \frac{1}{8} \cdot (1 - x_1x_2 +  y_1y_2 +  w[x_1 - x_2][y_1 + y_2] - x_1y_1x_2y_2)
\end{align*}
meaning if we exchange the values of $x_1$ and $x_2$ then these probabilities swap values.
The corresponding possibilities for the likelihood values are
$$
\left\{\begin{array}{l}
        (1-x_1)(1+x_2)(1+w)(1+y_1)(1+y_2) \\
        (1+x_1)(1-x_2)(1-w)(1+y_1)(1+y_2) \\
        (1-x_1)(1+x_2)(1-w)(1-y_1)(1-y_2) \\
        (1+x_1)(1-x_2)(1+w)(1-y_1)(1-y_2)
\end{array}\right.
$$
for site split $\{1\}$ and
$$
\left\{\begin{array}{l}
        (1+x_1)(1-x_2)(1+w)(1+y_1)(1+y_2) \\
        (1-x_1)(1+x_2)(1-w)(1+y_1)(1+y_2) \\
        (1+x_1)(1-x_2)(1-w)(1-y_1)(1-y_2) \\
        (1-x_1)(1+x_2)(1+w)(1-y_1)(1-y_2)
\end{array}\right.
$$
for site split $\{3\}$, which also both swap values when $x_1$ and $x_2$ are exchanged.

The same can be done for the splits $\{2\}$ and $\{1,2,3\}$ by exchanging $y_1$ and $y_2$ as well as $\{1,2\}$ and $\{1,3\}$ by exchanging both $x_1$ with $x_2$ and $y_1$ with $y_2$.
The split $\{1,3\}$ is unchanged by exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$.
In summary, exchanging $x_1$ and $x_2$ does not change the value of $\ell$, and thus $x_1=x_2$ at the maximum.
The analogous statement holds for $y_1$ and $y_2$.
The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
    \hline
$\siteSplit_j$   &$P(\siteSplitRV=\siteSplit_j \mid \tau_1,t)$&$P(\siteSplitRV=\siteSplit_j \mid \tau_2,t)$\\
    \hline
    $\emptyset$    &$1+x'^2+y'^2+4x'y'w'+x'^2y'^2$&$1+2x'y'+2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{1\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{2\}$        &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    $\{3\}$        &$1-x'^2+y'^2-x'^2y'^2$&$1-x'^2w'+y'^2w'-x'^2y'^2$\\
    $\{1,2\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1+2x'y'-2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,3\}$      &$1+x'^2+y'^2-4x'y'w'+x'^2y'^2$&$1-2x'y'-2x'y'w'+x'^2w'+y'^2w'+x'^2y'^2$\\
    $\{2,3\}$      &$1-x'^2-y'^2+x'^2y'^2$&$1-2x'y'+2x'y'w'-x'^2w'-y'^2w'+x'^2y'^2$\\
    $\{1,2,3\}$    &$1+x'^2-y'^2-x'^2y'^2$&$1+x'^2w'-y'^2w'-x'^2y'^2$\\
    \hline
\end{tabular}
\caption{Site pattern probabilities.
All values are multiplied by $1/8$.}
\label{tab:sitepatprob}
\end{table}

\begin{table}
\centering
\begin{tabular}{|l|ll|}
    \multicolumn{3}{c}{$\tau=\tau_1$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$&
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')^2$\\
     $\{1\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
     $\{2\}$    &
    $\emptyset$&
    $(1+x')^2   (1+w')(1+y')(1-y')$\\
     $\{3\}$    &
    $\emptyset$&
    $(1+x')(1-x')(1+w')(1+y')^2$\\
    $\{1,2\}$  &
    $\{\emptyset,\{1,2\}\}$&
    $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,3\}$  &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    %EM Shouldn't this be \{1\}\\ ? Or perhaps the ordering of internal nodes is not what I thought?
                    \{1\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1+w')(1-y')^2
                \end{array}$\\
    $\{2,3\}$  &
                $\{\emptyset,\{1,2\}\}$&
                $(1+x')(1-x')(1+w')(1+y')(1-y')$\\
    $\{1,2,3\}$&
                $\{1,2\}$&
                $(1+x')^2   (1+w')(1+y')(1-y')$\\
    \hline
    \multicolumn{3}{c}{$\tau=\tau_2$}\\
    \hline
    $\siteSplit_j$    & $\ancestralSplitPartition_j(\tau, t)$ & $P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j,\tau,t)$\\
    \hline
    $\emptyset$       &$\emptyset$&$(1+x')^2   (1+w')(1+y')^2$\\
    $\{1\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                        (1+x')(1-x')(1+w')(1+y')^2\\
                        (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
      $\{2\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')^2   (1+w')(1+y')(1-y')\\
                    (1+x')(1-x')(1-w')(1+y')^2
                    \end{array}$\\
      $\{3\}$          &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')^2\\
                    (1+x')^2   (1-w')(1+y')(1-y')
                    \end{array}$\\
     $\{1,2\}$         &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1+x')^2   (1-w')(1+y')^2
                    \end{array}$\\
     $\{1,3\}$         &
    $\left\{\begin{array}{l}
                    \emptyset\\
                    \{\{1\},\{2\}\}\\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1-x')^2   (1+w')(1+y')^2\\
                    (1+x')(1-x')(1-w')(1+y')(1-y')\\
                    (1+x')^2   (1+w')(1-y')^2
                    \end{array}$\\
      $\{2,3\}$        &
    $\left\{\begin{array}{l}
                    \{\emptyset,\{1,2\}\}\\
                    \{1\}\\
                    \{2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1+w')(1+y')(1-y')\\
                    (1-x')^2   (1-w')(1+y')^2\\
                    (1+x')^2   (1-w')(1-y')^2
                    \end{array}$\\
     $\{1,2,3\}$       &

                    \{2\} \\
                    \{1,2\}
                \end{array}\right.$&
    $\begin{array}{l}
                    (1+x')(1-x')(1-w')(1+y')^2 \\
                    (1+x')^2   (1+w')(1+y')(1-y')
                    \end{array}$\\
    \hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and ancestral state partitions.
All values multiplied by $1/32$.
Likelihoods with multiple entries have maxima determined by unknown branch length parameters.
See Appendix for full calculations.}
\label{tab:likelihoods}
\end{table}

@@

@@

%EM Afraid I don't see how all of the p_emptyset terms are bounded above here.

@@

The other two possible ancestral state splits for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
$$
a_{1}' = a_{1}-2p_{13},
$$
$$
a_{2}' = a_{2}+2p_{13},
$$
$$
a_{3}' = a_{3}-2p_{13},
$$
$$
a_{4}' = a_{4}+2p_{13}.
$$
Then,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
%EM Sorry to be slow, but could you provide a little more detail on why this follows? Is there some convexity of entropy argument going on here?
%Not sure which part is confusing, but:
    %1) a_1 log(1+x) + a_2 log(1-x) is maximized at x = (a_1 - a_2) / (a_1 + a_2).
    %The below equation follows by rewriting 1+x and 1-x for the above x.
    %2) a_1' and a_2' are just defined by rewriting the likelihood. There are now two fewer log(1+x) terms and two more log(1-x) terms since we replaced a (1+x)^2 in the likelihood with a (1-x)^2. That's why we have a -2p_{13} and a +2p_{13} respectively.
    %Happy to clarify somewhere in the text but not sure what part would most need exposition.
%EM Ha! Still working on understanding. It seems to me that the statement is equivalent to saying
%EM a_{1} \log a_{1} + a_{2} \log a_{2} + (1-p_{13}) \log(1-p_{13}) + p_{13}\cdot\log(2p_{13}) = a_{1}' \log a_{1}' + a_{2}' \log a_{2}'
%EM am I thinking about this wrong? Is this what you are saying in your 2)?
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
\end{align*}
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{\emptyset}  \cdot\log(2) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.
